Thesis Title

### 2022-10-24 chapters/conclusion.tex
First draft of conclusion

\begin{itemize}
    \item Standardise the process of metagenomic analysis, allowing the comparison of results obtained across domains and stakeholders
    \item Develop computationally efficient and robust frameworks that allow scientists and/or medical experts with limited programming experience to rapidly and easily query the abundance of specific taxa and genes across the samples of interest, obtaining simple and intuitive reports. 
\end{itemize}

The work I hereby present aims to evaluate the use of bioinformatics methods for the analysis of metagenomic data to allow the rapid identification, virulence analysis and antimicrobial susceptibility prediction of pathogens with clinical relevance.

\ac{SMg} still presents as a promising methodology to obtain very fast results for the identification of pathogens and their virulence and resistance properties directly from samples, without the need for culture. 
To be accredited and used in clinical settings, this approach must be standardised and the statistical metrics used to analyse and report the data must be validated.

@@

Chapters \ref{ch:paper3} and \ref{ch:paper4} were strongly focused on the standardisation and reproducibility of the results obtained, with the employment of new technologies to do so, such as container software and workflow managers.

>>>>

The work I hereby present aims to evaluate the use of bioinformatics methods for the analysis of metagenomic data to allow the rapid identification, virulence analysis and antimicrobial susceptibility prediction of pathogens with clinical relevance. \ac{SMg} still presents as a promising methodology to obtain very fast results for the identification of pathogens and their virulence and resistance properties directly from samples, without the need for culture. 
To be accredited and used in clinical settings, this approach must be standardised and the statistical metrics used to analyse and report the data must be validated.

@@

A strong focus on  the standardisation and reproducibility of the results obtained, with the employment of new technologies to do so, such as container software and workflow managers, is of the uttermost necessity for the \ac{SMg} data analysis solutions. Transparency, scalability, and ease of installation are key-stones, regardless of the tools chosen. The solutions adopted throughout this work, such the use of docker, nextflow and conda, allied to clear and easy to follow documentation, aim to lower the barrier of entry when performing detailed analysis that are complex and computationally expensive in their nature. But most importantly, the production of intuitive, responsive and easy-to-follow reports, allowing the summary of key results, as well as the detailed exploration of the resulting data, by stakeholders, be it bioinformatic personnel or experts in the given area of expertise, represent the single most important contribution to lowering the barrier between who produces the data and who has the capacity to make informed decisions based on that data.

As public health laboratories expand their genomic sequencing and bioinformatics capacity for the surveillance of different pathogens, labs must carry out robust validation, training, and optimisation of wet- and dry-lab procedures. Despite this richness in genomic information, the same is not observed for the contextual information that accompanies it. This contextual information, composed of metadata  Standardisation of harmonisable, publicly available community standards is ultimately what will allow the transmission of information through various stakeholders, containing the minimum information required for it to be understandable and actionable across domains.

\section{Future perspectives}

Despite the results presented in this work, the future of \ac{SMg} still looks promising. Regardless, for it to pass from the last resort solution when traditional methods fail to a standard in clinical microbiology, some issues still need to be addressed.

One of the biggest hinders to the application of \ac{SMg} in the clinic is the significant cost of negative results, allied with a very high sensitivity but not as high specificity, being very prone to contamination leading to false positive results. Therefore, part of the work to improve this technology passes through the improvement in the bench. Host depletion steps still have a long way to produce satisfactory results for diagnosis purposes, without altering the proportions of organisms in a community. Although traditional methods require a very significant amount of man-hours, sequencing requires a different, more specialised set of skills to obtain quality results. Additionally, the time gained in the results attainment is offset by the increased cost.

Regarding bioinformatic analysis of \ac{SMg} data, the lack of golden standards presents the biggest hindrance. But as new methods keep being developed, fuelled by the extreme and exciting interest of the community for this methodology, and more importantly, as current methods continue to be validated, and their results openly shared, in a faindable and FAIRer way, we slowly but surely walk towards the standardisation of methodologies.

I'm as excited with the potential of \ac{SMg} for clinical microbioly, both in the diagnostics and public health and surveillance branches, as I was when I first started this project. The potentials are staggering, albeit currently it's still very much an experimental and/or last resort approach. As methodologies continue to improve, sequencing will become ubiquitous in every diagnostics and public health laboratory. The transition between academic applications to being a viable solution in hospitals and health reference laboratories might be slow, or at least slower than what was once expected, but it's an inevitability that we should not only be prepared for but also excited about.
